NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2017-12-28 Manuscript/manuscriptPS.tex
Add the data by M. Javanainen for MacRog POPS and update figures and manuscript
Updates and reorganization of the manuscript

As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
surfactants. The changes with the addition of neutral lipids is significantly smaller.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

@@

%\section{Methods}

%\subsection{Molecular dynamics simulations}

\todo{When we have more data for Ca binding to PS containing bilayers, the discussion will be updated and PG results moved
to other manuscript.}
Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPGPS.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure}

\subsection{Effect of Ca$2+$ binding to PS headgroup}

Also the experimental order parameters for PS and PG headgroups
as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
\todo{These should be compared to simulations for potential structural interpretation of the changes.}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
  \caption{\label{PSPGchangesWITHCaCl}
    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
  }
  \todo{Get the small concentration data from the inserts}
\end{figure}
\begin{table*}[!p]

@@

\begin{minipage}[t]{\textwidth}

@@

\end{minipage}

@@

\section{Measurements of order parameter sign}

Fig. \ref{PShgSIGNS} summarizes the experimental results on the order parameter sign
measurement for POPS sample. The experimental protocol is the same used in Ref. \citenum{ferreira16}.
In (a) you see the headgroup region of the INEPT spectrum where alpha and beta are
identified. In (b) you have the R-PDLF slices for alpha and beta where you see one single
splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot be
calculated. On the bottom you have the S-DROSS slices of these two carbons. The grey lines show a
random collection of slices from noise such that it gets clear what is significant. The S-DROSS
slice for beta clearly shows that the order parameter is negative. The slice for alpha shows that
the higher order parameter is positive and suggests that the smaller order parameter is negative
(from the deviation towards negative values in the longer t1 times).
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
  \caption{\label{PShgSIGNS}
    Experimental results for sign measurement for POPS sample
  }
\end{figure}

The results updated with SIMPSON simulations for the SDROSS profiles
are shown in Fig. \ref{PShgSIGNSsimpson}. The value for the smaller
alpha order parameter is taken from Fig 3 in Ref. \citenum{roux91},
because resolution in 13C NMR experiments was nor high enough to determine
numerical value for this. The plots in Fig. \ref{PShgSIGNSsimpson} (c) show
the following. The error bars and points are the experimental SDROSS data.
The thick lines are SIMPSON simulations. The simulations were done by using
the order parameter for beta equal to -0.12 and for alpha one order parameter
equal to 0.09 and the other equal to -0.02 (black) or 0.02 (grey).
Since the black lines agree with experimental data, we conlude that
the order parameters for $\beta$ carbon are -0.12 and for $\alpha$
order parameters are 0.09 and -0.02.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PShgSIGNSsimpson.pdf}
  \caption{\label{PShgSIGNSsimpson}
    Experimental results for sign measurement for POPS sample
  }
\end{figure}

>>>>

@@

\section{Methods}

\subsection{Solid state NMR experiments}
The experimental protocol is the same used in Ref. \citenum{ferreira16}.

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data was collected with Open Collaboration method.
The simulated systems are listed in Table \ref{IONsystems}.
The simulation details are in SI or in references in the table.
\begin{table*}[!htb]

@@

%begin{minipage}[t]{\textwidth}

@@

%\end{minipage}

@@

\section{Headgroup and glycerol backbone order parameters measured from POPS lipid bilayer}

Figs. \ref{PShgSIGNSsimpson} and \ref{glycerolNMR} summarize the experimental NMR results for POPS
bilayer sample.

\begin{figure}[!htb]
  \centering
  \includegraphics[width=6.0cm]{../Figs/PShgSIGNSsimpson.pdf}
  \caption{\label{PShgSIGNSsimpson}
    (a) The headgroup region of the INEPT spectrum where alpha and beta are
    identified.
    (b) The R-PDLF slices for alpha and beta showing one single
    splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
    of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot
    resolved with the available resolution.
    (c) Points with error bars are the experimental SDROSS data.
    The thick lines are SIMPSON simulations.
    The S-DROSS slice for beta clearly shows that the order parameter is negative,
    which is confirmed by SIMPSON simulations using the order parameter value of -0.12.
    The S-DROSS slice for alpha suggests that the higher order parameter is positive
    and the deviation towards negative values in the longer t1 times suggests
    that the smaller order parameter is negative.
    This is confirmed by SIMPSON simulation using value of 0.09
    for the larger alpha order parameter and the value of -0.02 for
    smaller (black curve). The value for the smaller
    alpha order parameter for SIMPSON calculation was taken from Fig 3 in Ref. \citenum{roux91},
    because resolution in 13C NMR experiments was nor high enough to determine
    numerical value for this. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller order parameter gave did not agree with experiments (dashed grey),
    confirming the interpretation that the smaller order parameter is negative.
  }
  \todo{Maybe we should combine this with \ref{glycerolNMR}}
\end{figure}

\begin{figure}[!htb]
  \centering
  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
  \caption{\label{glycerolNMR}
    R-PDFL slices for gycerol backbone carbons.
  }
  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
  \todo{What are the top figures actually?}
\end{figure}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
%  \caption{\label{PShgSIGNS}
%    Experimental results for sign measurement for POPS sample
%  }
%\end{figure}

\todo{When we have more data for Ca binding to PS containing bilayers, the discussion will be updated and PG results moved
to other manuscript.}
Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge
\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure*}

%\subsection{Effect of Ca$2+$ binding to PS headgroup}

Also the experimental order parameters for PS and PG headgroups
as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
\todo{These should be compared to simulations for potential structural interpretation of the changes.}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
%  \caption{\label{PSPGchangesWITHCaCl}
%    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
%  }
%  \todo{Get the small concentration data from the inserts}
%\end{figure}

\section{PC lipid headgroup response to different mixtures in experiments}
As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
surfactants. The changes with the addition of neutral lipids is significantly smaller.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}
